---
title: "RtEstimation EpiNow_PL"
author: "Phoebe Lu"
date: "`r Sys.Date()`"
output: html_document
---

### Load libraries
```{r load-libraries}
library(EpiNow2)
library(dplyr)
library(lubridate)
```

### Pull data
-   reference_date = hospitalization date
-   `ma_nn`: Massachusetts state data with no delay (all report_date=reference_date)
-   `ma_dn`: Massachusetts state data with delay (report date != reference_date, or in other words, individuals are not reported on the same day they are hospitalized)
-   All datasets subset for a 90-day period (10/3/25-12/31/25) due to EpiNow2 computational constraints
```{r data_pull}
#Read in data
#MA no noise
data_path <- paste0("data/Test data/MA/")
ma_nn <- read.csv(paste0(data_path, "MA_NoNoise.csv")) %>%
  rename("date"="reference_date",
         "confirm"="count") %>%
  dplyr::select(date, confirm) %>%
  mutate(date=as.Date(date, format="%Y-%m-%d")) %>%
  filter(date >= "2025-10-03" & date <= "2025-12-31")

#MA delayed noise
ma_dn <- read.csv(paste0(data_path, "MA_DelayNoise.csv"))  %>%
  mutate(reference_date=as.Date(reference_date, format="%Y-%m-%d"),
         report_date=as.Date(report_date, format="%Y-%m-%d")) %>%
  dplyr::filter(report_date <= "2025-12-31") %>%
  group_by(reference_date) %>%
  mutate(confirm=sum(count)) %>%
  rename("date"="reference_date") %>%
  dplyr::select(date, confirm) %>%
  dplyr::filter(row_number()==1)
  
```

### Define distributions
-   Since EpiEstim and Epifilter use a combined gamma distribution for IP and GT (mean=4, sd=sqrt(8)), we deconvolve to separate gamma distributions for EpiNow2, which requires separate arguments
-   Assume data reporting delay is mean=1, sd=0.1
```{r define-dist}
#Generation time
generation_time <- dist_spec(mean=2, sd=2, distribution="gamma", max=10)

#Incubation period
incubation_period <- dist_spec(mean=2, sd=2, distribution="gamma", max=10)

#Data reporting delay
data_report_delay <- dist_spec(mean=1, sd=0.1, distribution="gamma", max=10)
```


### Run EpiNow2
```{r epinow-run}
#Set parameters
rw_value <- 1
# Set horizon
horizon <- 21
# Empty call turns truncation correction off
epinow2_truncation_module <- trunc_opts()
# Basic settings
n_chains <- 2
n_samples <- 2000
n_warmup <- 600
seed <- 123
# CFA gp settings
epinow2_gp <- EpiNow2::gp_opts(alpha_sd = 0.0075)

epinow_run <- EpiNow2::epinow(reported_cases=ma_dn, # Input dataset
                              rt=EpiNow2::rt_opts(prior = list(mean=1, sd=0.2), # Prior for Rt
                                                  rw=rw_value),
                              truncation=epinow2_truncation_module,
                              # This parameter sets the prior on the sd of the gaussian process
                              # We are using this value to penalize large jumps in Rt from one day to the next
                              gp = epinow2_gp,
                              generation_time = generation_time_opts(generation_time), #generation time dist estimation
                              delays=delay_opts(incubation_period+data_report_delay), 
                              # These options tell the model to use a Poisson error structure, and to adjust for day-of-week effects
                              obs = EpiNow2::obs_opts(
                                family="poisson", 
                                week_effect=TRUE),
                              stan=EpiNow2::stan_opts(
                                cores=4,
                                chains=n_chains,
                                seed=seed,
                                control=list(
                                  adapt_delta=0.99,
                                  max_treedepth=12
                                ),
                                samples=n_samples, 
                                warmup=n_warmup
                              ),
                              horizon=horizon, #7 day forecast
                              CrIs=c(0.05, 0.95),
                              return_output=T,
                              verbose=interactive()) 
```

### Save output
```{r save-output}
#Obtain Rt estimates results summary
model_summary <- epinow_run$estimates$summarised
#Subset for 90 days of Rt estimates and the relevant columns
model_fnl <- model_summary %>%
  head(90) %>%
  dplyr::select(date, median, lower_95, upper_95)
#Write to file
write.csv(model_fnl, "analysis/rt-estimation/EpiNow/MA_Rt_estimates_EpiNow_DelayNoise.csv")
```


